---
title: 'Data Filtering'
author: '<b>ImmunoMind</b> – improving design of T-cell therapies using multi-omics and AI. Research and biopharma partnerships, more details: <a href="https://immunomind.io">immunomind.io</a>'
date: "support@immunomind.io"
output:
  html_document:
    fig_height: 8
    fig_width: 10
    theme: spacelab
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---


<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Gene usage analysis}
%\VignettePackage{immunarch}
-->



```{r setup, include=FALSE, echo=FALSE}
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
# knitr::opts_chunk$set(optipng = '-o7')

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(fig.width = 12)
knitr::opts_chunk$set(fig.height = 6)

library(immunarch)
```

# Data filtering
In many research cases, you could want to filter your data by metadata, clonotypes parameters or genes, so for this purpose, you can use the `repFilter` function. 

## Methods for filtering data
`repFilter` has 3 parameters: `.method`, `.query` and `.match`.

Due to the ambiguity of ways to extract required data, `repFilter` has the following methods:

- `.method = "by.meta"` - filters data using information from meta data.

- `.method = "by.repertoire" or "by.rep"` - filters data using information about repertoire statistics.

- `.method = "by.clonotype" or "by.cl"` - filters data using information about clonotype sequences.


Parameter `.query` is a list specifying the condition for filtering. Elements in the list are conditions, names of elements - columns chosen for filtering. Names in elements should not be repeated in `.query`. If you need to write a complex condition, you can call the function multiple times.

Parameter `.match` denotes the type of match between condition and value to filter: 

 - `exact` - looks for exact match.
 - `substring`- looks for substring.
 - `startswith` - looks for string starting with the some pattern

Parameter `.match` can be applied to both by.meta and by.clonotype


Load the package into the R environment:
```{r }
library(immunarch)
```

For testing purposes we will use `scdata` and `immdata` datasets from **Immunarch** library. Load them by executing the following command:

```{r}
data(immdata)
```

Look at meta data from `immdata` datasets:

```{r}
#look at the metadata
immdata$meta
#look at samples name in data
names(immdata$data)
```


## How to write conditions for filtering

### Method `by.meta`

#### Example 1
Use  `include` and `exclude` options to select a subset of samples or repertoires that match given filter criterion. These options accept strings.

Let's filter your data by metadata `immdata` datasets when *Status* is *C* and look at new meta data:
```{r example-1.1}
repFilter(immdata, .method = "by.meta", .query = list(Status = include("C")))$meta
```

Note that we filter both metadata and data:

```{r example-1.1.1}
# filtered data
names(repFilter(immdata, .method = "by.meta", .query = list(Status = include("C")))$data)
```

Filter out samples where *Lane* is not A:

```{r example-1.2}
repFilter(immdata, .method = "by.meta", .query = list(Lane = exclude("A")))$meta
names(repFilter(immdata, .method = "by.meta", .query = list(Lane = exclude("A")))$data)
```

Filter for samples where *Lane* is *B* or *C*:

```{r example-1.3}
repFilter(immdata, .method = "by.meta", .query = list(Lane = include("B", "C")))$meta
names(repFilter(immdata, .method = "by.meta", .query = list(Lane = include("B", "C")))$data)
```

Filter out samples where  *Lane* is not *A* and not *C*:

```{r example-1.4}
repFilter(immdata, .method = "by.meta", .query = list(Lane = exclude("A", "C")))$meta
names(repFilter(immdata, .method = "by.meta", .query = list(Lane = exclude("A", "C")))$data)
```

#### Example 2
Use  `interval`, `lessthan`(from minus infinity to your value) or `morethan`(from your value to plus infinity) to define interval values in the filter statement. These options accept float. `lessthan` and `morethan` do not include borders and `interval` includes the left border and excludes the right border.

Filter for samples where *Age* is lower than *23*:
```{r example-2.1}
repFilter(immdata, .method = "by.meta", .query = list(Age = lessthan(23)))$meta
names(repFilter(immdata, .method = "by.meta", .query = list(Age = lessthan(23)))$data)
```

Filter for samples where *Age* is upper than *15*:

```{r example-2.2}
repFilter(immdata, .method = "by.meta", .query = list(Age = morethan(15)))$meta
names(repFilter(immdata, .method = "by.meta", .query = list(Age = morethan(15)))$data)
```

Filter for samples where *Age* is between *15* and *23*:

```{r example-2.3}
repFilter(immdata, .method = "by.meta", .query = list(Age = interval(15, 23)))$meta
names(repFilter(immdata, .method = "by.meta", .query = list(Age = interval(15, 23)))$data)
```

You can also use **multiple conditions**. In this case, the function returns values that matches both of this conditions (logical AND operator)

Filter for samples where *Age* is between *15* and *23* and *Lane* is *B* :

```{r example-2.4}
repFilter(immdata, .method = "by.meta", .query = list(Age = interval(15, 23), Lane = include('B')))$meta
names(repFilter(immdata, .method = "by.meta", .query = list(Age = interval(15, 23), Lane = include('B')))$data)
```

Filter for samples where *Age* is between *15* and *23* and *Lane* is *A* or *B*:

```{r example-2.5}
repFilter(immdata, .method = "by.meta", .query = list(Age = interval(15, 23), Lane = include('A', 'B')))$meta
names(repFilter(immdata, .method = "by.meta", .query = list(Age = interval(15, 23), Lane = include('A', 'B')))$data)
```

### Method `by.repertoire` (short alias is `by.rep`)

#### Example 3

Filter for repertoires containing more than 60000 clonotypes:

```{r example-3.1}
repFilter(immdata, .method = "by.repertoire", .query = list(n_clonotypes = morethan(6000)))$meta
repFilter(immdata, .method = "by.rep", .query = list(n_clonotypes = morethan(6000)))$meta # Works both with by.rep and by.repertoire
```

Be careful, filtering `by.repertoire` or `by.rep` could also change a number of repertoires(samples) in your data: 

```{r example-3.1.1}
names(repFilter(immdata, .method = "by.repertoire", .query = list(n_clonotypes = morethan(6000)))$data)
```

Filter for repertoires containing less than 6000 clonotypes:

```{r example-3.2}
repFilter(immdata, .method = "by.repertoire", .query = list(n_clonotypes = lessthan(6000)))$meta
repFilter(immdata, .method = "by.rep", .query = list(n_clonotypes = lessthan(6000)))$meta # Works both with by.rep and by.repertoire
names(repFilter(immdata, .method = "by.repertoire", .query = list(n_clonotypes = lessthan(6000)))$data)
```

### Method `by.clonotype` (short alias is`by.cl`)

#### Example 4

Filter out all noncoding clonotypes from `immdata`. As you see, `immdata` dataset doesn't contain any noncoding clonotypes:

```{r example-4.1}
repFilter(immdata, .method = "by.clonotype", .query = list(CDR3.aa = exclude("partial", "out_of_frame")))$meta
```

Note that filtering `by.clonotype` or `by.cl` works within repertoire(sample). Repertoire(sample) could be removed from your data only in case if all clonotypes in sample do not satisfy the condition: 

```{r example-4.1.1}
names(repFilter(immdata, .method = "by.clonotype", .query = list(CDR3.aa = exclude("partial", "out_of_frame")))$data)
```

Filter out clonotypes that have only one clone:

```{r example-4.2}
repFilter(immdata, .method = "by.clonotype", .query = list(Clones = morethan(1)))$meta
names(repFilter(immdata, .method = "by.clonotype", .query = list(Clones = morethan(1)))$data)
```

#### Example 5

In method `by.clonotype` or `by.cl`, there is an extra argument `.match`. The `.match` argument can has the following values: 
 - `exact` - looks for exact match in gene names
 - `substring`- looks for substring in gene names
 - `startswith` - looks for gene names starting with the same pattern
 
Filter out all clonotypes within samples with V gene 'TRBV1' or 'TRGV11'

```{r example-5.1, results = 'hide'}
repFilter(immdata, .method = "by.clonotype", .query = list(V.name = exclude("TRBV1", "TRGV11")), .match="exact")
```

Filter out all clonotypes within samples where V gene name contains substrings 'TRBV1' or 'TRGV11'
```{r example-5.2, results = 'hide'}
repFilter(immdata, .method = "by.clonotype", .query = list(V.name = exclude("TRBV1", "TRGV11")), .match="substring")

```

Filter out all clonotypes within samples where V gene name starts with 'TRBV1' or 'TRGV11'
```{r example-5.3, results = 'hide'}
repFilter(immdata, .method = "by.clonotype", .query = list(V.name = exclude("TRBV1", "TRGV11")), .match="startswith")
```

## Using `repFilter` function for repertoire data analysis 

We will use `scdata` datasets from **Immunarch** library. Load them by executing the following command:
```{r}
data(scdata)
```


Look at meta data from `scdata` datasets:

```{r scadata}
#look at the metadata
scdata$meta
#look at samples name in data
names(scdata$data)
```

`repFiter` can also work with single-cell data containing not only `meta` and 'data`, but also extra information, for example, about clusters:

```{r, results = 'hide'}
repFilter(scdata, .method = "by.clonotype", .query = list(CDR3.aa = exclude("partial", "out_of_frame")))
```

Create a new dataset with cluster-specific immune repertoires (for more information see *Single-cell* tutorials):
```{r}
scdata_cl <- select_clusters(scdata, scdata$bc_cluster, "Cluster")
scdata_cl$meta
```

And compare the number of clonotypes between clusters:

```{r clonotypes, warnings = F}
vis(repExplore(scdata_cl$data, .method = "volume"))
```

Compare J gene usage between three clusters:

```{r geneUsage,  warning=F}
sc_active <- geneUsage(repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Activ')))$data, "hs.trbj", .norm = T)
p1 <- vis(sc_active)
p1

sc_memory <- geneUsage(repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Memory')))$data, "hs.trbj", .norm = T)
p2 <- vis(sc_memory)
p2

sc_naive <- geneUsage(repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Naive')))$data, "hs.trbj", .norm = T)
p3 <- vis(sc_naive)
p3

```

Compare gene usage of IGHJ4 between three clusters:
```{r IGHV3, warning=F}
scdata_active <- repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Activ')))
scdata_active <- repFilter(scdata_active, .method = "by.cl", .query = list(J.name = include('IGHJ4')), .match = 'substring')
sc_active <- geneUsage(scdata_active$data, "hs.trbj", .norm = T)
p1 <- vis(sc_active)

scdata_memory <- repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Memory')))
scdata_memory <- repFilter(scdata_memory, .method = "by.cl", .query = list(J.name = include('IGHJ4')), .match = 'substring')
sc_memory <- geneUsage(scdata_memory$data, "hs.trbj", .norm = T)
p2 <- vis(sc_memory)

scdata_naive <- repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Naive')))
scdata_naive <- repFilter(scdata_naive, .method = "by.cl", .query = list(J.name = include('IGHJ4')), .match = 'substring')
sc_naive <- geneUsage(scdata_naive$data, "hs.trbj", .norm = T)
p3 <- vis(sc_naive)

p1 + p2 + p3
```

Look at the coding clonotypes in each cluster. Note that there aren't any noncoding clonotypes in the datasets:
```{r ORF, warning=F}
scdata_active <- repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Activ')))
scdata_active <- repFilter(scdata_active, .method = "by.clonotype", .query = list(CDR3.aa = exclude("partial", "out_of_frame")))
exp_vol <- repExplore(scdata_active$data, .method = "volume")
p1 <- vis(exp_vol)

exp_vol <- repExplore(repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Activ')))$data, .method = "volume")
p2 <- vis(exp_vol)
p1+p2


scdata_memory <- repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Memory')))
scdata_memory <- repFilter(scdata_memory, .method = "by.clonotype", .query = list(CDR3.aa = exclude("partial", "out_of_frame")))
exp_vol <- repExplore(scdata_memory$data, .method = "volume")
p1 <- vis(exp_vol)

exp_vol <- repExplore(repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Memory')))$data, .method = "volume")
p2 <- vis(exp_vol)
p1+p2

scdata_naive <- repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Naive')))
scdata_naive <- repFilter(scdata_naive, .method = "by.clonotype", .query = list(CDR3.aa = exclude("partial", "out_of_frame")))
exp_vol <- repExplore(scdata_naive$data, .method = "volume")
p1 <- vis(exp_vol)

exp_vol <- repExplore(repFilter(scdata_cl, .method = "by.meta", .query = list(Cluster = include('Naive')))$data, .method = "volume")
p2 <- vis(exp_vol)
p1+p2
```