---
title: 'Data Filtering'
author: '<b>ImmunoMind</b> – improving design of T-cell therapies using multi-omics and AI. Research and biopharma partnerships, more details: <a href="https://immunomind.io">immunomind.io</a>'
date: "support@immunomind.io"
output:
  html_document:
    fig_height: 8
    fig_width: 10
    theme: spacelab
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---


<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Gene usage analysis}
%\VignettePackage{immunarch}
-->



```{r setup, include=FALSE, echo=FALSE}
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
# knitr::opts_chunk$set(optipng = '-o7')

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(fig.width = 12)
knitr::opts_chunk$set(fig.height = 6)

library(immunarch)
```

# Data filtering
In many research cases, you want to filter your data by metadata, clonotypes parameters or genes, so for this purpose, you can use the `repFilter` function. 

## Methods for filtering data
`repFilter` has 2 parameters: `method ` and `query`.

Due to the ambiguity of ways to extract required data, `repFilter` has the following methods:

- `method = "by.meta"` - filters data using information from meta data.

- `method = "by.repertoire" or "by.rep"` - filters data using information about repertoire statistics.

- `method = "by.clonotype" or "by.cl"` - filters data using information about clonotype sequences.


Parameter `query` is a list specifying the condition for filtering. Elements in the list are conditions, names of elements - columns chosen for filtering. Names in elements should not be repeated in query. If you need to write a complex condition, you can call the function multiple times.

Load the package into the R environment:
```{r }
library(immunarch)
```

For testing purposes we will use `scdata` and `immdata` datasets from **Immunarch** library. Load them by executing the following command:

```{r}
data(immdata)
```

Look at meta data from `immdata` datasets:

```{r}
#look at the metadata
immdata$meta
#look at samples name in data
names(immdata$data)
```


## How to write conditions for filtering

### Method `by.meta`

#### Example 1
Use  `include` and `exclude` options to select a subset of samples or repertoires that match given filter criterion. These options accept strings.

Let's filter your data by metadata `immdata` datasets when *Status* is *C* and look at new meta data:
```{r example-1.1}
# method = "by.meta", query = list(Status = include("C"))
immdata$meta %>% filter(Status == 'C')
```
This way we have filtered both metadata and data:

```{r example-1.1.1}
# filtered data
names(immdata$data)
```

Filter out samples where *Lane* is not A:

```{r example-1.2}
# method = "by.meta", query = list(Lane = exclude("A"))
immdata$meta %>% filter(Status != 'C')
names(immdata$data)
```

Filter for samples where *Lane* is *B* or *C*:

```{r example-1.3}
# method = "by.meta", query = list(Lane = include("B", "C"))
immdata$meta %>% filter(Lane == 'C' | Lane == 'B')
names(immdata$data)
```

Filter out samples where  *Lane* is not *A* and not *C*

```{r example-1.4}
# method = "by.meta", query = list(Lane = exclude("B", "C"))
immdata$meta %>% filter(Lane != 'C' & Lane != 'A')
names(immdata$data)
```

#### Example 2
Use  `interval`, `lessthan`(from minus infinity to your value) or `morethan`(from your value to plus infinity) to define interval values in the filter statement. These options accept float.

Filter for samples where *Age* is lower than *23*:
```{r example-2.1}
#method = "by.meta", query = list(Age = lessthan(23)) 
immdata$meta %>% filter(Age < 23)
names(immdata$data)
```

Filter for samples where *Age* is upper than *15*:

```{r example-2.2}
# method = "by.meta", query = list(Age = morethan(15))
immdata$meta %>% filter(Age > 15)
names(immdata$data)
```

Filter for samples where *Age* is between *15* and *23*:

```{r example-2.3}
# method = "by.meta", query = list(Age = interval(15, 23))
immdata$meta %>% filter(Age > 15 & Age < 23)
names(immdata$data)
```

You can also use **multiple conditions**. In this case, the function returns values that matches both of this conditions (logical AND operator)

Filter for samples where *Age* is between *15* and *23* and *Lane* is *B*:

```{r example-2.4}
# method = "by.meta", query = list(Age = interval(15, 23), Lane = include('B'))
immdata$meta %>% filter(Age > 15 & Age < 23 & Lane == 'B')
names(immdata$data)
```

### Method `by.repertoire` (short alias is `by.rep`)

#### Example 3

Filter for repertoires  containing more than 60000 clonotypes:

```{r example-3.1}
#method = "by.repertoire", query = list(n_clonotypes = morethan(6000))
#method = "by.rep", query = list(n_clonotypes = morethan(6000)) # Works both with by.rep and by.repertoire
```

Note, filtering `by.repertoire` or `by.rep` could change a number of repertoires or samples in your data: 

```{r example-3.1.1}
names(immdata$data)
```

Filter for repertoires containing less than 4000 clonotypes:

```{r example-3.2}
#method = "by.repertoire", query = list(n_clonotypes = lessthan(4000))
#method = "by.rep", query = list(n_clonotypes = lessthan(4000)) # Works both with by.rep and by.repertoire
names(immdata$data)
```

### Method `by.clonotype` or `by.cl`

#### Example 4

Filter for all noncoding clonotypes from `immdata` . As can be seen, `immdata` datasets doesn't contain any noncoding clonotypes:

```{r example-4.1}
#method = "by.clonotype", query = list(CDR3.aa = exclude("partial", "out_of_frame"))

```

Note, filtering `by.clonotype` or `by.cl` works within repertoire or sample. Repertoire or sample could be removed from your data only in case if all clonotypes in sample do not satisfy the condition: 

```{r example-4.1.1}
#method = "by.clonotype", query = list(CDR3.aa = exclude("partial", "out_of_frame"))
names(immdata$data)
```

Filter out clonotypes with 1 clones from `immdate`:

```{r example-4.2}
#method = "by.clonotype", query = list(Clones = morethan(2))
names(immdata$data)
```

#### Example 5

In method `by.clonotype` or `by.cl`, there are an extra argument `match`. The argument has the following method: 
 - `exact` - looks for exact match in gene names
 - `by substring`- looks for substring in gene names
 - `by start` - looks for gene names starting with the some pattern
 
Filter out all clonotypes within samples with V gene 'TRBV1' or 'TRGV11'

```{r example-5.1}
#method = "by.clonotype", query = list(V.gene = exclude("TRBV1", "TRGV11")), match="exact"
```

Filter out all clonotypes within samples where V gene name contains substrings 'TRBV1' or 'TRGV11'
```{r example-5.2}
#method = "by.clonotype", query = list(V.gene = exclude("TRBV1", "TRGV11")), match="substring"

```

Filter out all clonotypes within samples where V gene name starts with 'TRBV1' or 'TRGV11'
```{r example-5.3}
#method = "by.clonotype", query = list(V.gene = exclude("TRBV1", "TRGV11")), match="startswith"
```

## Using `repFilter` function for repertoire data analysis 

We will use `scdata` datasets from **Immunarch** library. Load them by executing the following command:
```{r}
data(scdata)
```


Look at meta data from `scdata` datasets and create a new dataset with cluster-specific immune repertoires (for more information see *Single-cell* tutorials):

```{r}
scdata_cl <- select_clusters(scdata, scdata$bc_cluster, "Cluster")
scdata_cl$meta
```

And compare the number of clonotypes between clusters:

```{r clonotypes, warnings = F}
vis(repExplore(scdata_cl$data, .method = "volume"))
```

Compare J gene usage between 3 cluster:

```{r geneUsage,  warning=F}
# method = "by.meta", query = list(Cluster = include('Activ'))
sc_active <- geneUsage(scdata_cl[1], "hs.trbj", .norm = T)
p1 <- vis(sc_active)
p1
# method = "by.meta", query = list(Cluster = include('Memory'))
sc_memory <- geneUsage(scdata_cl[1], "hs.trbj", .norm = T)
p2 <- vis(sc_memory)
p2
# method = "by.meta", query = list(Cluster = include('Naive'))
sc_naive <- geneUsage(scdata_cl[1], "hs.trbj", .norm = T)
p3 <- vis(sc_naive)
p3

```

Compare gene usage of IGHV3 between 3 cluster:
```{r IGHV3, warning=F}
# method = "by.meta", query = list(Cluster = include('Activ'))
# method = "by.gene", query = list(J = include('IGHJ4'))
sc_active <- geneUsage(scdata_cl$data$flu_Activ %>% filter(grepl('IGHV3',V.name)), "hs.trbj", .norm = T)
p1 <- vis(sc_active)
# method = "by.meta", query = list(Cluster = include('Memory'))
# method = "by.gene", query = list(J = include('IGHJ4'))
sc_memory <- geneUsage(scdata_cl$data$flu_Memory %>% filter(grepl('IGHV3',V.name)), "hs.trbj", .norm = T)
p2 <- vis(sc_memory)
# method = "by.meta", query = list(Cluster = include('Naive'))
# method = "by.gene", query = list(J = include('IGHJ4'))
sc_naive <- geneUsage(scdata_cl$data$flu_Naive %>% filter(grepl('IGHV3',V.name)), "hs.trbj", .norm = T)
p3 <- vis(sc_naive)
p1
p2
p3

```

Look at the coding and nonconding clonotypes in each cluster. As can be easily seen, there is no noncoding clonotypes in the datasets:
```{r ORF, warning=F}
# method = "by.meta", query = list(Cluster = include('Active'))
#method = "by.clonotype", query = list(CDR3.aa = exclude("partial", "out_of_frame"))
scdata_cod <- coding(scdata_cl$data$flu_Activ)
exp_vol <- repExplore(scdata_cod, .method = "volume")
p1 <- vis(exp_vol)
# method = "by.meta", query = list(Cluster = include('Active'))
#method = "by.clonotype", query = list(CDR3.aa = exclude("in_frame", "partial"))
scdata_cod <- noncoding(scdata_cl$data$flu_Activ)
exp_vol <- repExplore(scdata_cod, .method = "volume")
p2 <- vis(exp_vol)
p1+p2


# method = "by.meta", query = list(Cluster = include('Memory'))
#method = "by.clonotype", query = list(CDR3.aa = exclude("partial", "out_of_frame"))
scdata_cod <- coding(scdata_cl$data$flu_Memory)
exp_vol <- repExplore(scdata_cod, .method = "volume")
p1 <- vis(exp_vol)
# method = "by.meta", query = list(Cluster = include('Memory'))
#method = "by.clonotype", query = list(CDR3.aa = exclude("in_frame", "partial"))
scdata_cod <- noncoding(scdata_cl$data$flu_Memory)
exp_vol <- repExplore(scdata_cod, .method = "volume")
p2 <- vis(exp_vol)
p1+p2

# method = "by.meta", query = list(Cluster = include('Naive'))
#method = "by.clonotype", query = list(CDR3.aa = exclude("partial", "out_of_frame"))
scdata_cod <- coding(scdata_cl$data$flu_Naive)
exp_vol <- repExplore(scdata_cod , .method = "volume")
p1 <- vis(exp_vol)
# method = "by.meta", query = list(Cluster = include('Naive'))
#method = "by.clonotype", query = list(CDR3.aa = exclude("in_frame", "partial"))
scdata_cod <- noncoding(scdata_cl$data$flu_Naive)
exp_vol <- repExplore(scdata_cod, .method = "volume")
p2 <- vis(exp_vol)
p1+p2
```