---
title: "Single-cell and paired chain data: scTCRseq and scBCRseq exploration"
author: '<b>ImmunoMind</b> – improving design of T-cell therapies using multi-omics and AI. Research and biopharma partnerships, more details: <a href="https://immunomind.io">immunomind.io</a>'
date: "support@immunomind.io"
output:
  html_document:
    fig_height: 8
    fig_width: 10
    theme: spacelab
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---


<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Working with data in immunarch}
%\VignettePackage{immunarch}
-->



```{r setup, include=FALSE, echo=FALSE}
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
# knitr::opts_chunk$set(optipng = '-o7')

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(fig.width = 18)
knitr::opts_chunk$set(fig.height = 12)

library(immunarch)
data(scdata)
```


# Executive Summary

> This is a vignette dedicated to provide an overview on how to work with single-cell paired chain data in `immunarch`

> Single-cell support is currently in the development version. In order to access it, you need to install the latest development version of the package  by executing the following command:

```r
install.packages("devtools"); devtools::install_github("immunomind/immunarch", ref="dev")
```

> To read paired chain data into `immunarch` use the `repLoad` function with `.mode = "paired"`. Currently we support 10X Genomics only.

> To subset immune repertoires by specific barcodes use the `select_barcodes` function. Output of `Seurat::Idents()` as a barcode vector works.

> To create cluster-specific and patient-specific datasets using barcodes from the output of `Seurat::Idents()` use the `select_clusters` function.


# Use the data packaged with `immunarch`

Load the package into the R enviroment:
```{r}
library(immunarch)
```

For testing purposes we attached a new paired chain dataset to `immunarch`. Load it by executing the following command:

```{r}
data(scdata)
```

# Load the paired chain data

To load your own datasets, use the `repLoad` function. Currently we implemented paired chain data support for 10X Genomics data only. A working example of loading datasets into R:

```{r}
file_path <- paste0(system.file(package = "immunarch"), "/extdata/sc/flu.csv.gz")
igdata <- repLoad(file_path, .mode = "paired")

igdata$meta

head(igdata$data[[1]][c(1:7, 16, 17)])
```

# Subset by barcodes

To subset the data by barcodes, use the `select_barcodes` function.

```{r}
barcodes <- c("AGTAGTCAGTGTACTC-1", "GGCGACTGTACCGAGA-1", "TTGAACGGTCACCTAA-1")

new_df <- select_barcodes(scdata$data[[1]], barcodes)

new_df
```

## Patient-specific datasets

To create a new dataset with cluster-specific immune repertoires, use the `select_clusters` function:

```{r}
scdata_pat <- select_clusters(scdata, scdata$bc_patient, "Patient")

names(scdata_pat$data)

scdata_pat$meta
```

## Cluster-specific datasets

To create a new dataset with cluster-specific immune repertoires, use the `select_clusters` function. You can apply this function after you created patient-specific datasets to get patient-specific cell cluster-specific immune repertoires, e.g., a Memory B Cell repertoire for a specific patient:

```{r}
scdata_cl <- select_clusters(scdata_pat, scdata$bc_cluster, "Cluster")

names(scdata_cl$data)

scdata_cl$meta
```

# Explore and compute statistics

Most functions will work out-of-the-box with paired chain data.

```{r}
p1 <- repOverlap(scdata_cl$data) %>% vis()
p2 <- repDiversity(scdata_cl$data) %>% vis()

target <- c("CARAGYLRGFDYW;CQQYGSSPLTF", "CARATSFYYFHHW;CTSYTTRTTLIF", "CARDLSRGDYFPYFSYHMNVW;CQSDDTANHVIF", "CARGFDTNAFDIW;CTAWDDSLSGVVF", "CTREDYW;CMQTIQLRTF")
p3 <- trackClonotypes(scdata_cl$data, target, .col = "aa") %>% vis()

(p1 + p2) / p3
```

Several functions may work incorrectly with paired chain data in this release of `immunarch`. Let us know via [GitHub Issues](http://github.com/immunomind/immunarch/issues)!
