---
title: "Make your plots publication-ready with fixVis"
author: '<b>ImmunoMind</b> – improving design of T-cell therapies using multi-omics and AI. Research and biopharma partnerships, more details: <a href="https://immunomind.io">immunomind.io</a>'
date: "support@immunomind.io"
output:
  html_document:
    fig_height: 8
    fig_width: 10
    theme: spacelab
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---


<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Make your plots publication-ready with fixVis}
%\VignettePackage{immunarch}
-->



```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(fig.width = 12)
knitr::opts_chunk$set(fig.height = 6)

library(immunarch)
```

In order to make the visualization process easier, we developed a Shiny-based tool called FixVis. It is distributed with the `immunarch` package. To run FixVis just apply it to any ggplot2-based plot, including those from the `immunarch`'s `vis` function. It works perfectly for every plot. FixVis can work only with one plot at a time.


Here is an example of how to run FixVis with  `vis` function from `immunarch`. First, the plot is generated by `vis` function from `immunarch` and then FixVis is applied:

```{r fixvis1, eval=F}
data(immdata)
gu <- geneUsage(immdata$data)
p <- vis(gu)
fixVis(p)
```

If you want to test it, run `fixVis` with no arguments; it will use the diamonds data for visualisation:

```{r fixvis2, eval=F}
fixVis()
```

For your convenience we published a FixVis instances that are available online [here](https://immunomind.shinyapps.io/fixvis/). Also, a quick overview of FixVis is embedded below:

```{r fixvis3, eval=T}
knitr::include_app("https://immunomind.shinyapps.io/fixvis/", height = "800px")
```
