---
title: 'Make your plots publication-ready with fixVis'
author: '<a href="https://immunomind.io">ImmunoMind</a>'
date: "support@immunomind.io"
output:
  html_document:
    fig_height: 8
    fig_width: 10
    theme: spacelab
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---


<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Make your plots publication-ready with fixVis}
%\VignettePackage{immunarch}
-->



```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(fig.width = 12)
knitr::opts_chunk$set(fig.height = 6)

library(immunarch)
```

In order to make visualisation process easier, we developed a Shiny-based tool called FixVis. The FixVis is distributed with `immunarch` package. To run FixVis just apply it to any ggplot2-based plot, including those from the `immunarch`'s `vis` function. It works perfectly for every plot, except those arranged in grid by `grid.arrange` due to the complexities surrounding `grid.arrange`.

```{r fixvis1, eval=F}
data(immdata)
gu = geneUsage(immdata$data)
p = vis(gu)
fixVis(p)
```

If you want to test it, run `fixVis` with no arguments; it will use the diamonds data for visualisation:

```{r fixvis2, eval=F}
fixVis()
```

For convenience we published a FixVis instances that are available online [here](https://immunomind.shinyapps.io/fixvis/). Also a FixVis instance for quick overview is embedded in the cell below:

```{r fixvis3, eval=T}
knitr::include_app("https://immunomind.shinyapps.io/fixvis/", height = "800px")
```
