---
title: "Data loading"
author: '<b>ImmunoMind</b> – improving design of T-cell therapies using multi-omics and AI. Research and biopharma partnerships, more details: <a href="https://immunomind.io">immunomind.io</a>'
date: "support@immunomind.io"
output:
  html_document:
    fig_height: 8
    fig_width: 10
    theme: spacelab
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---


<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Working with data in immunarch}
%\VignettePackage{immunarch}
-->



```{r setup, include=FALSE, echo=FALSE}
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
# knitr::opts_chunk$set(optipng = '-o7')

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(fig.width = 12)
knitr::opts_chunk$set(fig.height = 6)

library(immunarch)
# source("../R/testing.R")
# immdata = load_test_data()
data(immdata)
```


# Input / output
The package provides several IO functions:

- `repLoad` - to load the repertoires, having compatible format.

- `repSave` - to save changes and to write the repertoire data to a file in a specific format (`immunarch`, VDJtools).

`repLoad` detects the input file format automatically. `immunarch` currently support the following immune repertoire data formats:

- `"immunarch"` - current software tool, in case you forgot it :)

- `"immunoseq"` - https://www.immunoseq.com

- `"mitcr"` - https://github.com/milaboratory/mitcr

- `"mixcr"` - https://github.com/milaboratory/mixcr

- `"migec"` - http://migec.readthedocs.io/en/latest/

- `"migmap"` - https://github.com/mikessh/migmap

- `"tcr"` - https://imminfo.github.io/tcr/

- `"vdjtools"` - https://vdjtools-doc.readthedocs.io/en/master/

- `"imgt"` - http://www.imgt.org/HighV-QUEST/

- `"airr"` - http://docs.airr-community.org/en/latest/datarep/overview.html

- `"10x"` - https://support.10xgenomics.com/single-cell-vdj/software/pipelines/latest/output/annotation

- `"archer"` - ArcherDX clonotype tables.

- `"imseq"` - http://www.imtools.org/

- `"rtcr"` - https://github.com/uubram/RTCR

- `"vidjil"` - http://www.vidjil.org/

Please contact us if there are more file formats you want to be supported.

For parsing IgBLAST results process the data with MigMap first.

You can load the data either from a single file, a list of repertoire file paths or from a folder with repertoire files. 

**Working with your files**


If you have your files, you should just specify a path to your file or to a folder with files. Then load data using `repLoad`:

```{r, eval=F}
#path argument is a path to the folder with your file or files including metadata file also(if you have one).
immdata <- repLoad(path)
```

**Working with sample files**


You could find a folder with example files [here](https://github.com/immunomind/immunarch/tree/master/inst/extdata/io) and use it to test data loading without your own files. 

If you are not familiar with file paths, you can download example data to your working directory.
You can obtain you working directory with `getwd()` command
```{r, eval=F}
getwd()

You could also download all files to the `'example'` folder in your working directory and load all of them by passing folder name to repLoad function in a quotes:

```{r, eval=F}
immdata <- repLoad('example')
```

Also, example data are already downloaded with `immunarch` package. You can load all sample files using the following command:

```{r, eval=F}
#path to the folder with example data
file_path = paste0(system.file(package="immunarch"), "/extdata/io/")
immdata <- repLoad(file_path)
```

In other cases you may want to provide a metadata file and locate it in the folder. It is necessary to name it exactly "metadata.txt".

```{r, eval=F}
# For instance you have a following structure in your folder:
# >_ ls
# immunoseq1.txt
# immunoseq2.txt
# immunoseq3.txt
# metadata.txt
```

With the metadata `repLoad` will create a list in the environment with 2 elements, namely `data` and `meta`. All the data will be accessible simply from `immdata$data`.

Otherwise `repLoad` will create a dummy metadata file with only sample names.

```{r, eval=F}
# To load the whole folder with every file in it type:
file_path = paste0(system.file(package="immunarch"), "/extdata/io/")
immdata <- repLoad(file_path)
print(names(immdata))

# In order to do that your folder must contain metadata file named
# exactly "metadata.txt".

# In R, when you load your data:
# > immdata <- repLoad("path/to/your/folder/")
# > names(immdata)
# [1] "data" "meta"

# Suppose you do not have "metadata.txt":
# > immdata <- repLoad("path/to/your/folder/")
# > names(immdata)
# [1] "data" "meta"

```

Dummy metadata data frame look like this:

```{r}
as_tibble(data.frame(Sample = c("immunoseq1", "immunoseq2", "immunoseq3"), stringsAsFactors = F))
```

The metadata file "metadata.txt" has to be tab delimited file with first column named "Sample" and any number of additional columns with arbitrary names. The first column should contain base names of files without extensions in your folder.

| **Sample** |**Sex**|**Age**|**Status**|
|:----------:|:-----:|:-----:|:--------:|
|immunoseq\_1|M      |1      |C         |
|immunoseq\_2|M      |2      |C         |
|immunoseq\_3|F      |3      |A         |

In order to import data from the external databases you have to create a connection to this database and then load the data. Make sure that the table format in your database matches the `immunarch`'s format.

To illustrate the use of external database, here is an example demonstrating data loading to the local MonetDB database:
```{r, eval=F}
# Your list of repertoires in immunarch's format
DATA
# Metadata data frame
META

# Create a temporary directory
dbdir = tempdir()

# Create a DBI connection to MonetDB in the temporary directory.
con = DBI::dbConnect(MonetDBLite::MonetDBLite(), embedded = dbdir)

# Write each repertoire to MonetDB. Each table has corresponding name from the DATA
for (i in 1:length(DATA)) {
  DBI::dbWriteTable(con, names(DATA)[i], DATA[[i]], overwrite=TRUE)
}

# Create a source in the temporary directory with MonetDB
ms = MonetDBLite::src_monetdblite(dbdir = dbdir)
res_db = list()

# Load the data from MonetDB to dplyr tables
for (i in 1:length(DATA)) {
  res_db[[names(DATA)[i]]] = dplyr::tbl(ms, names(DATA)[i])
}

# Your data is ready to use
list(data = res_db, meta = META)
```


`immunarch` is compatible with following sources:

- R data frames (for common applications)

- R data tables (for speed, although requires lots of RAM)

- MonetDB-like databases that support both DBI and dplyr (best choice, although you have to be a bit familiar with dplyr)

- Apache Spark (if you are expierienced and comfortable with it)

# Basic data manipulations with dplyr and immunarch
You can find the introduction to `dplyr` here: https://CRAN.R-project.org/package=dplyr/vignettes/dplyr.html

## Get the most abundant clonotypes
The function returns the most abundant clonotypes for the given repertoire:
```{r basic-data}
top(immdata$data[[1]])
```

## Filter functional / non-functional / in-frame / out-of-frame clonotypes
Conveniently, functions are vectorised over the list of data frames; and `coding(immdata$data)` in the example below returns a list of data frames with coding sequences:
```{r, eval=FALSE}
coding(immdata$data[[1]])
```
The next one operates in a similar fashion:
```{r, eval=FALSE}
noncoding(immdata$data[[1]])
```
Now, the computation of the number of filtered sequences is straightforward:
```{r, eval=FALSE}
nrow(inframes(immdata$data[[1]]))
```
And for the out-of-frame clonotypes:
```{r, eval=FALSE}
nrow(outofframes(immdata$data[[1]]))
```

## Get subset of clonotypes with a specific V gene
It is simple to subset data frame according to labels in the specified index. In the example the resulting data frame contains only records with 'TRBV10-1' V gene:
```{r}
filter(immdata$data[[1]], V.name == 'TRBV10-1')
```

# Downsampling

```{r}
ds = repSample(immdata$data, "downsample", 100)
sapply(ds, nrow)
```

```{r}
ds = repSample(immdata$data, "sample", .n = 10)
sapply(ds, nrow)
```


# Immunarch data format
`immunarch` comes with its own data format, including tab-delimited columns that can be specified as follows:

 - "Clones" - count or number of barcodes (events, UMIs) or reads;

 - "Proportion" - proportion of barcodes (events, UMIs) or reads;

 - "CDR3.nt" - CDR3 nucleotide sequence;

 - "CDR3.aa" - CDR3 amino acid sequence;

 - "V.name" - names of aligned Variable gene segments;

 - "D.name" - names of aligned Diversity gene segments or NA;

 - "J.name" - names of aligned Joining gene segments;

 - "V.end" - last positions of aligned V gene segments (1-based);

 - "D.start" - positions of D'5 end of aligned D gene segments (1-based);

 - "D.end" - positions of D'3 end of aligned D gene segments (1-based);

 - "J.start" - first positions of aligned J gene segments (1-based);

 - "VJ.ins" - number of inserted nucleotides (N-nucleotides) at V-J junction (-1 for receptors with VDJ recombination);

 - "VD.ins" - number of inserted nucleotides (N-nucleotides) at V-D junction (-1 for receptors with VJ recombination);

 - "DJ.ins" - number of inserted nucleotides (N-nucleotides) at D-J junction (-1 for receptors with VJ recombination);

 - "Sequence" - full nucleotide sequence.

<!---
## Additional functions to help with dplyr

- `todf` - convert the input to R data frame.

- `todt` - convert the input to R data table.

- `db_add` - add a new column to the existing database.

- `db_rem` - remove a specific column from the existing database.

- `db_write` - write / rewrite a database table.

With these functions you can write the subset of your repertoires to new tables with `db_write`, after you have filtered out the noncoding sequences, for instance.

-->
