---
title: "Comprehensive analysis of single-cell immune repertoire data using `immunarch` and `immundata`"
author: "Vadim I. Nazarov"
date: "`r Sys.Date()`"
format: 
  html:
    code-overflow: wrap
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    highlight-style: github
    code-line-numbers: false
    code-copy: true
    toc: true
    number-sections: true
    # theme: none
    # minimal: true
vignette: >
  %\VignetteIndexEntry{Comprehensive analysis of single-cell immune repertoire data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette provides a comprehensive guide on how to analyse single-chain and paired-chain Adaptive Immune Receptor Repertoire (AIRR) data coming from scRNAseq (scVDJseq / scTCRseq / scBCRseq). This tutorils guides you from the very basics to the more advanced steps. The aim is to have a great starting point for both beginners in AIRR data analysis and experts who want specificaly learn `immunarch` and `immundata`.

We will use data from the study for patient no. 6. The data will have three samples - tumor, blood, and healthy tissue surrounding the tumor. The core idea of the study is to ...

To ask a question, report a bug, or take a look at the source code, following the GitHub links for [immunarch](https://github.com/immunomind/immunarch) and [immundata](https://github.com/immunomind/immundata).

To use ChatGPT for interactive helper, use [AIRR-GPT](https://chatgpt.com/g/g-67fe54e302348191814e4ef2efb3bad5-airr-gpt) - a GPT Store model with uploaded documentation for both `immunarch` and `immundata`.

### Migration from v0.9 to v1.0

This is a section for user who already use immunarch. If you are a new user, you can safely skip it and move directly to the **Setup** section, where we will install `immunarch` (together!) and start analysing *your* immune repertoire data.

#### High-level overview of v1.0 changes

There are several important changes in the v1.0 version in comparison to v0.9.x versions.

1) working with data

2) API changes - backwards compatability

**Watch for Deprecation Warnings**
    -   If you see console messages about “deprecated” or “will be removed in v1.1.0,” update your code ASAP.\

3) moving packages from Depends to Suggests

3) working with visualisatiobns - hard to make publication ready - https://github.com/CerrenRichards/ggplot2-for-publications

Take a look at themes here: https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html and https://emilhvitfeldt.github.io/r-color-palettes/ 

For more information and updates, please see the guide here: ...

To list updates, run the function in console

#### If you want to work with `immunarch` v0.9 **only**

To freeze the latest stable version of `immunarch` before `1.0`, you could provide the exact version:

 - **pak** --- `pak::pkg_install("immunarch@=0.9.1")`
 
 - **install.packages** --- `install.packages("https://cran.r-project.org/src/contrib/Archive/immunarch/immunarch_0.9.1.tar.gz", repos = NULL, type = "source")`
 
 - **conda** --- `mamba install -c conda-forge r-immunarch=0.9.1`

#### New data processing backend --- `immundata`

...

#### Motivation behind the changes

...


#### Concluding thoughts on migration

...

---

## Setup

### Pre-requisites

...

```{r}
#| eval: false
install.packages("pak", repos = sprintf("https://r-lib.github.io/p/pak/stable/%s/%s/%s", .Platform$pkgType, R.Version()$os, R.Version()$arch))

pak::pkg_install("immunomind/immunarch")
pak::pkg_install("Seurat")
pak::pkg_install("BiocFileCache")
pak::pkg_install("ggthemes")
pak::pkg_install("ggsci")
```

### Load necessary packages and setup

...

```{r}
#| label: load-immunarch
#| eval: !expr requireNamespace("Seurat", quietly = TRUE) && requireNamespace("ggthemes", quietly = TRUE)
#| echo: true
#| message: false
#| warning: false

library("immunarch")
library("Seurat") 

theme_set(ggthemes::theme_few())
```

---

## Quick overview

### Typical workflow

analyse -> visualise -> write back to Seurat -> visualise Seurat

---

## Working with data

### Load the prepackaged dataset

First, we get paths to the metadata table and load it. Metadata is a tab-separate table with rows per sample or file, and columns for different per-sample feature such as "Tissue" or "Donor" or "Immunotherapy response".

```{r}
#| label: load-dataset-metadata

md_file <- system.file("extdata/single_cell", "metadata.tsv", package = "immundata")
md_file

md_table <- immundata::read_metadata(md_file)
md_table
```

The metadata table contains the column "Tissue" and we will use it to split data into repertoires. But this is only the beginning - read more to learn how to split data by annotated cell clusters from scRNAseq data and computate statistics on those newly defined repertoires - on-the-fly and without reloading the whole dataset into the environment again.

Next, we get the file paths to the data packaged with `immundata`.

```{r}
#| label: load-dataset-files
#| 
inp_files <- paste0(system.file("extdata/single_cell", "", package = "immundata"), "/*.csv.gz")
inp_files
```

TODO: EXPLAIN MORE WHAT THE RECEPTOR IS IN IMMUNDATA MODEL

Then, we are ready to load AIRR data. We define a logical unit of analysis - a receptor - as a `"cdr3+v_call"` and will use only the `"TRB"` chain:

```{r}
#| label: load-dataset-schema
#| eval: true

schema <- make_receptor_schema(features = c("cdr3", "v_call"), chains = c("TRB"))
schema
```

If you want to use other receptor schemas, you can use the `make_receptor_schema()` function to create schemas and pass them to `read_repertoires()`. The dataset we use for this tutorial can support other schemas as well, including the paired-chain data:

```{r}
#| label: load-dataset-schema-examples
#| eval: true

# Paired-chain
schema <- make_receptor_schema(features = c("cdr3", "v_call"), chains = c("TRA", "TRB"))
schema

# Single-chain with "cdr3" only
schema <- make_receptor_schema(features = c("cdr3"), chains = c("TRA"))
schema

# The most strict receptor definition
schema <- make_receptor_schema(features = c("cdr3", "v_call", "j_call"), chains = c("TRA", "TRB"))
schema
```

Use the `read_repertoires()` function to load the data. There are quite many parameters, so let's briefly describe them:

- `path` - ...

- `schema` - ...

- `metadata` - ...

```{r}
#| label: load-dataset-airr
#| eval: true

schema <- make_receptor_schema(features = c("cdr3", "v_call"), chains = c("TRB"))

idata <- read_repertoires(path = inp_files, schema = schema, metadata = md_table, barcode_col = "barcode", locus_col = "locus", umi_col = "umis", preprocess = make_default_preprocessing("10x"), repertoire_schema = "Tissue")

# idata
```

The output `idata` is `ImmunData` - a data structure from `immundata` package that holds the AIRR dataset. Think `AnnData` or `SeuratObject`, but for AIRR data. It is not a simple data frame or a list of data frames, and in the next subsection we will talk about how to operate on it.

We have two optional steps left, and those are needed to load single-cell transcriptomics data. If you analyse bulk data, you obviously don't need to do this.

We start with reading a cell metadata - a convenient human-readable table with T cell cluster annotations. Please note that in most cases, you will use `SeuratObject` or `AnnData` directly to exctact cluter labels from the object to annotate `ImmunData`. Read the corresponding documentation on how to extract cluster labels from single-cell transcriptomic and other tools.

```{r}
#| label: load-dataset-cells
#| eval: true

cells_file <- system.file("extdata/single_cell", "cells.tsv.gz", package = "immundata")
cells_file

cells <- readr::read_tsv(cells_file)
cells
```

And finally, we load the gene expression data to Seurat. The matrix object takes several hundreds of megabytes, so we use cached file to avoid re-downloading the file next time we run the tutorial. Obviously, this wouldn't work without the web connection.

The `cells` object contains also precomputed coordinates for UMAP to reduce the computations time for this tutorial. In real world, you would run the typical processing pipeline for scRNAseq data and get UMAP coordinates on the `SeuratObject`. To simulate this, we will add the UMAP coordinates to the created object in the code block below.

```{r}
#| label: load-dataset-seurat
#| eval: true

url <- "https://zenodo.org/records/15604205/files/l6data.rds?download=1"
dest <- BiocFileCache::bfcrpath(BiocFileCache::BiocFileCache(ask = FALSE), url)

mat <- readr::read_rds(dest)

sdata <- CreateSeuratObject(counts = mat)

embeddings <- as.matrix(cells[c("UMAP_1", "UMAP_2")])
rownames(embeddings) <- cells$barcode

umap_dr <- CreateDimReducObject(embeddings = embeddings, key = "UMAP_", assay = DefaultAssay(sdata))

sdata[["umap"]] <- umap_dr

Idents(sdata) <- setNames(cells$ident, cells$barcode)

DimPlot(sdata, reduction = "umap", order = sort(unique(cells$ident), decreasing = TRUE), label = TRUE, alpha = .5)
```


#### How to read other types of data - bulk, single-chain, etc.

Note that the code below is for demonstration purposes only. It is not meant to run as a part of this tutorial.

1. Bulk

You won't need `barcode_col`, `locus_col`, `umi_col`. You can optionally pass `count_col` to have receptors counts.

```{r}
#| label: load-dataset-types-bulk
#| eval: false

idata <- read_repertoires(path = inp_files, schema = schema, metadata = md_table, count_col = "counts", preprocess = make_default_preprocessing("airr"), repertoire_schema = "Tissue")
```

2. Single-cell

All columns are needed.

```{r}
#| label: load-dataset-types-singlecell
#| eval: false

idata <- read_repertoires(path = inp_files, schema = schema, metadata = md_table, barcode_col = "barcode", locus_col = "locus", umi_col = "umis", preprocess = make_default_preprocessing("10x"), repertoire_schema = "Tissue")
```

3. AIRR or 10XGenomics

`immunarch` aims to support the AIRR-C standard only. However, there are a couple of format-specific processing utilities. Mind the locus column name and the identifier we pass the to `make_default_preprocessing` function.

```{r}
#| label: load-dataset-types-10x
#| eval: false

idata <- read_repertoires(path = inp_files, schema = schema, metadata = md_table, barcode_col = "barcode", locus_col = "locus", umi_col = "umis", preprocess = make_default_preprocessing("airr"), repertoire_schema = "Tissue")

idata <- read_repertoires(path = inp_files, schema = schema, metadata = md_table, barcode_col = "barcode", locus_col = "chain", umi_col = "umis", preprocess = make_default_preprocessing("10x"), repertoire_schema = "Tissue")
```

4. Pass file paths via metadata table - metadata_file_col

Pass `"<metadata>"` 

```{r}
#| label: load-dataset-types-meta
#| eval: false

idata <- read_repertoires(path = "<metadata>", schema = schema, metadata = md_table, barcode_col = "barcode", locus_col = "locus", umi_col = "umis", preprocess = make_default_preprocessing("10x"), repertoire_schema = "Tissue")
```

---

## Note on materialization

`ImmunData` object works on out-of-memory datasets the same way it works on in-memory datasets. It saves necessary computations and runs them only when the data is needed, e.g., for visualization. "Materialization" refers to running all the computations and getting this "real" dataset. This behaviour is not very familiar to people. Therefore, if you encounter this error:

```
Error: Materialization is disabled, use collect() or as_tibble() to materialize.
```

it means that the data is not materialized yet. If the data is not huge, e.g., it is some statistics from `immunarch` functions, you can safely run `immunarch_function() |> collect()` to immediately collect this data. If the dataset is huge, e.g., filtered `ImmunData` object, you usually don't need to materialize it so please think twice before doing this.

---

## Analyse immune repertoires - individually and in groups


### Explore key statistics of immune repertoires


#### Number of receptors and barcodes for samples

```{r}
#| label: repertoire-stats-sample-plot
#| eval: true
#| fig-width: 11
idata_stats <- airr_stats_chains(idata)

idata_stats

p1 <- ggplot(idata_stats, aes(x = Tissue, y = n_barcodes, fill = Tissue)) + geom_col() + ggtitle("No. barcodes per sample") + ggsci::scale_fill_locuszoom()
p2 <- ggplot(idata_stats, aes(x = Tissue, y = n_receptors, fill = Tissue)) + geom_col() + ggtitle("No. receptors per sample") + ggsci::scale_fill_locuszoom()

p1 + p2
```


#### Number of receptors and barcodes for cell clusters

Okay, that's great, but what if we want to have a detailed view on how clusters are represented? Currently, we have three repertoires in the `idata` coming from the different tissues:

```{r}
#| label: repertoire-stats-show-repertoires
#| eval: true

idata$repertoires
```

Ideally, we want to explore each cell subpopulation in each tissue. For this, we will use powerful annotation mechanism from `immundata`. `idata` and `cells` have the same cell barcodes, so we can use them to annotate immune receptors with cluster label information, and then compute the same statistics on cluster level.

```{r}
#| label: repertoire-stats-cluster
#| eval: true

annot <- tibble(Cluster = as.character(Idents(sdata)), barcode = names(Idents(sdata)))

annot

idata <- annotate_barcodes(idata, annot, "barcode")
idata <- agg_repertoires(idata, c("Tissue", "Cluster"))

idata$repertoires
```

And now, let's plot recompute statistics and plot them, but applied to "new" repertoires, which are now defined by a pair of columns - a tissue name "Tissue" and a cluster label "Cluster". 

```{r}
#| label: repertoire-stats-cluster-plot
#| eval: true
#| fig-width: 11
idata_stats <- airr_stats_chains(idata)

idata_stats

p1 <- ggplot(idata_stats, aes(x = Tissue, y = n_barcodes, fill = Cluster)) + geom_col() + ggtitle("No. barcodes per sample")
p2 <- ggplot(idata_stats, aes(x = Tissue, y = n_receptors, fill = Cluster)) + geom_col() + ggtitle("No. receptors per sample")

p1 + p2
```

That's a lot of information! Let's group by cluster to understand the differences between tissues.

```{r}
#| label: repertoire-stats-cluster-plot-patients
#| eval: true
#| fig-width: 11

ggplot(idata_stats, aes(x = Cluster, y = n_receptors, fill = Tissue)) + geom_col(position = "dodge") + ggtitle("No. receptors per sample") + ggsci::scale_fill_locuszoom() + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

Lots of NAs. This is because we took only "TRB" chain, and the tumor receptors are going crazy, they have lots of receptors per one barcode.

If we reload the data with paired chain data, we will see another picture, as `immunarch` will filter out receptors additional receptors per barcode and leave only one pair per barcode. Do you want to do it? Consider this your homework! You just need to define a paired-chain schema for receptors (see one of the previous sections) and run pretty much the same code without any changes - that's the power of decoupling data processing to `immundata` and data analysis to `immunarch`. The latter pretty much doesn't care about the former, it just "asks" for receptors to exists - in any shape or form.

We see some expected yet interesting patterns like more Tregs in the tumor site or more Trm in the surrounding normal tissue. Our data is not normalized so we can't really compare how big is the difference. We will normalize the data by the overall number of receptors per sample, and then plot the same visualization.

```{r}
#| label: repertoire-stats-cluster-plot-norm
#| eval: true
#| fig-width: 11

idata_stats2 <- idata_stats |> mutate(.by = Tissue, TissueSize = sum(n_barcodes)) |> mutate(n_barcodes_div = n_barcodes / TissueSize)
idata_stats2

ggplot(idata_stats2, aes(x = Cluster, y = n_barcodes_div, fill = Tissue)) + geom_col(position = "dodge") + ggtitle("No. barcodes per sample") + ggsci::scale_fill_locuszoom() + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

#### Gene usage

```{r}
#| label: repertoire-gene-usage-plot
#| eval: true
#| fig-width: 11

# filter Trm and Treg
# compute stats
# join them
# visualise
airr_stats_genes(idata)
```


#### Length distribution


### Clonality

...


### Diversity

...


### Compute public receptor indices

...


#### Number of public receptors

...


#### Jaccard index

...


But what are those public receptors? The next section will help us extract them from the data.

---

## Discover and annotate immune receptors

This section is under construction as I currently develop the receptor-level analysis functionality. I plan to release the most commonly used functions in October-November 2025. Monitor my [LinkedIn](https://www.linkedin.com/in/vdnaz) and [GitHub](https://github.com/immunomind/immunarch/issues/432) for news and updates.

The planned topics are:

- use external databases to find CMV-specific receptors in the input repertoires

- track specific CMV-related receptors through different tissues and highlight them on single-cell UMAP

- analyse and plot counts of specific receptors vs. expression of genes of interest

- public repertoire analysis - discover receptors which are suspiciously overabundant in one groups of repertoires in contrast to another groups of repertoires (e.g., control vs condition to search for TIL or antigen-specific receptors)

- all of above but using levenshtein-based similarity instead of exact matches

---

## Analyse subgroups of receptors of interest

This is section is too under constructions. Planned content:

- discover receptors of interest, e.g., CMV-specific

- run analysis to compare sum/avg abundance of such receptors between groups to measure the level of signal

---

## Conclusion

No spoilers. :-)
